How to build an homogeneous language for heterogeneous platforms

### 2013-05-23 papers/gpce2013/gpce2013.tex
Taking shape

\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{graphicx}

@@

@@

This paper presents such a compiled embedded DSL allowing developers to write Web applications in a single language
which code fragments can be shared between client and server sides, and which is efficiently compiled to each side.
More precisely, we demonstrate the following features:

\begin{itemize}
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs.
\end{itemize}

@@

@@

As an illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL provides an
abstraction at the stage-level that is removed by optimization during the code generation.

@@

@@

\begin{figure}
\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
def vertexDom(v: Rep[Vertex]) =
    el("g", "class"->"vertex",
            "transform"->("translate("+v.x+","+v.y+")"))(
        el("rect", "width"->v.width, "height"->v.height)(),
        el("text", "width"->v.width, "height"->v.height)(
            v.content
        )
    )
\end{lstlisting}
\end{figure}

Listing \ref{forest-hello} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.

\subsubsection{Reuse the DOM definition DSL on server-side}

@@

For instance, the template written in listing \ref{forest-hello} produces the following Scala code usable on
server-side (the generated code for client-side is roughly equivalent to listing \ref{dom-api}):

\begin{lstlisting}
def vertexDom(v: Vertex) = {
  val x0 =
    <text width="{v.width}" height="{v.height}">
      {v.content}
    </text>
  val x1 = <rect width="{v.width}" height="{v.height}" />
  val x2 =
    <g class="vertex" transform="translate({v.x},{v.y})">
      {List(x0, x1)}
    </g>
  x2
}
\end{lstlisting}

We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
and Scala is managed by their respective code generators.

@@

@@

querySelectorAll, getElementsByTagName, getElementsByClassName, getElementById, querySelector

\section{Implementation?}

>>>>

\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{graphicx}

@@

(Mention the problem of using client-side libraries to get high-level abstractions: double runtime overhead --
download and execution)

@@

\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is a compiled embedded DSL defined in Scala that generates JavaScript
code. In this paper we present additional contributions on top of this DSL. We use staging (explain) to generate
efficient code for typical abstractions used for Web programming and to generate efficient specialized code for both
client and server sides for shared abstractions. More precisely, we demonstrate the following features:

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
 \item (Modules?)
\end{itemize}

@@

In a Web application, the user interface is defined by a HTML document which can be manipulated by the JavaScript
code to be updated. One typical operation consists in searching some “interesting” element in the document, in order
to extract its content, replace it or listen to user events fired on this element. The standard API provides several
function to search elements in a Web document according to their characteristics such as their attribute values.
Figure~\ref{selectors-api} (wrong reference number) summarizes the available functions and their differences.

\begin{figure}
\label{selectors-api}
\begin{tabular}{| l | p{3cm} |}
\hline
Function & Description \\
\hline
\code{querySelector(s)} & First element matching the CSS selector \code{s} \\
\hline
\code{getElementById(i)} & Element which attribute \code{id} equals to \code{i} \\
\hline
\code{querySelectorAll(s)} & All nodes matching the CSS selector \code{s} \\
\hline
\code{getElementsByTagName(n)} & All elements of type \code{n} \\
\hline
\code{getElementsByClassName(c)} & All elements which \code{class} attribute equals to \code{c} \\
\hline
\end{tabular}
\caption{Standard selectors API}
\end{figure}

The \code{querySelector} and \code{querySelectorAll} are the most general functions, the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery library (ref and stats) that provides just one function to search for
elements. Listing~\ref{jquery-vanilla-api} shows two equivalent snippets of code, the first one using the native
APIs (a) and the second one using jQuery (b).

\begin{figure}
\begin{lstlisting}[language=JavaScript]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\caption{(a)}
\begin{lstlisting}
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\caption{(b)}
\label{jquery-vanilla-api}
\end{figure}

jQuery provides an API that is simpler to master because it has less functions, but by doing so it can not benefit
from the performance of the browser’s implementation of specialized search functions (\code{getElementById},
\code{getElementsByTagName} and \code{getElementsByClassName}).

Using LMS we are able to provide a simple API (like jQuery) and to generate code using specialized functions (and
hence benefit from their performances), by analyzing the staged program. Listing~\ref{js-scala-selectors} shows how
to implement listing~\ref{jquery-vanilla-api} using js-scala. This listing generates a JavaScript program similar to
listing~\ref{jquery-vanilla-api} (a).

\begin{figure}
\label{js-scala-selectors}
\begin{lstlisting}
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\caption{Selectors in Js-Scala}
\end{figure}

Writing the program with js-scala instead of using a JavaScript library such as jQuery is interesting because the
abstraction for searching elements exist only in the initial program source code, not in the final JavaScript
program. It gives two advantages: (1) the execution of the final program is more performant because of the use of
specialized APIs, (2) the final program’s size is smaller because it does not need to include jQuery.

@@

As another illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL
provides an abstraction at the stage-level that is removed by optimization during the code generation.

@@

\begin{figure}
\label{dom-api}
\begin{lstlisting}[language=JavaScript]
var articleUi = function (article) {
  var div = document.createElement('div');
  div.setAttribute('class', 'article');
  var span = document.createElement('span');
  var name =
    document.createTextNode(article.name + ': ');
  span.appendChild(name);
  div.appendChild(span);
  var strong = document.createElement('strong');
  var price = document.createTextNode(article.price);
  strong.appendChild(price);
  div.appendChild(strong);
  return div
};
\end{lstlisting}

\begin{lstlisting}[language=HTML]
<div class=article>
  <span>French wine: </span>
  <strong>10</strong>
</div>
\end{lstlisting}
\caption{DOM}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[label=forest,caption=DOM definition DSL]
def articleUi(article: Rep[Article]) =
    el('div, 'class -> 'article)(
        el('span)(article.name + ": "),
        el('strong)(article.price)
    )
\end{lstlisting}
\end{figure}

Listing \ref{forest} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.

@@

For instance, the template written in listing \ref{forest} produces the following Scala code usable on
server-side:

\begin{lstlisting}
def articleUi(article: Article) = {
  val x0 = <span>{ article.name + ": " }</span>
  val x1 = <strong>{ article.price }</strong>
  val x2 =
    <div class="article">
      {List(x0, x1)}
    </div>
  x2
}
\end{lstlisting}

We are able to tackle the code sharing issues described in the introduction because of the embdedded nature
of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
and Scala is managed by their respective code generators.

@@

\section{Implementation}
\label{implementation}

@@

\begin{figure}
\label{selector-impl}
\begin{lstlisting}
def find(receiver: Exp[SelectorApi],
         selector: Exp[String]) =
  getConstIdentifier(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\caption{Selectors optimization}
\end{figure}
